---
title: "Speed Limits, Traffic Volume, and Fatal Crash Risk in North Carolina"
subtitle: "PLAN 372 Final Project"
author: "Liam MacKinnon"
date: ""
format:
  pdf:
    toc: false
    number-sections: true
    geometry: margin=1in
    fontsize: 12pt
    linestretch: 2
    latex-engine: xelatex
    keep-tex: false
execute:
  warning: false
  message: false
  echo: false
  cache: false
---

# Introduction and Motivation

Last year I was hit by a car while walking across a crosswalk. I was not seriously injured, but the experience made me a lot more aware of how dangerous roads can be for everyone on them, not just drivers. It also made me curious about what the data actually says about vehicle fatalities and what factors make crashes more deadly. That personal experience is what motivated me to look into traffic fatality patterns in North Carolina for this project.

Traffic fatalities are one of the most preventable public health problems in the United States. In North Carolina alone, hundreds of people die in motor vehicle crashes every year. Unlike a lot of public health issues, traffic deaths follow predictable patterns tied to road design, speed, and traffic volume, which means policy has a real opportunity to make a difference.

This report looks at the relationship between posted speed limits, traffic volume (measured as Annual Average Daily Traffic, or AADT), and fatal crash outcomes across North Carolina from 2018 through 2022. The main research question is: **Are posted speed limits and traffic volume associated with fatal crash outcomes in North Carolina, and does that relationship differ between urban and rural areas?**

The findings are directly relevant to North Carolina's Vision Zero initiative, which aims to eliminate all traffic fatalities on state roads. If speed limits are a strong predictor of crash severity, that supports lowering limits on dangerous corridors. If traffic volume is the bigger factor, the focus should shift toward infrastructure improvements on busy roads.

# Literature Review

The link between speed and crash severity is one of the most consistent findings in traffic safety research. Nilsson (2004) developed the "Power Model," which estimates that a 1% increase in average traffic speed leads to about a 2% increase in injury crashes and a 4% increase in fatal crashes. That nonlinear relationship has been validated across multiple countries and road types (Elvik et al., 2004). Farmer (2019) found supporting evidence in the U.S. context: when states raised speed limits on interstates between 1993 and 2017, fatality rates on those roads went up by about 8.5%.

Rural roads are a particular concern in North Carolina. Zwerling et al. (2005) found that rural residents face significantly higher traffic fatality rates than urban residents because of higher speed limits, longer emergency response times, and fewer safety features. North Carolina's rural counties, especially in the eastern coastal plain and the mountains, show some of the highest crash fatality rates in the state (NCDOT, 2023).

The relationship between traffic volume and crash severity is more complicated. Lord and Mannering (2010) found that crashes scale less than proportionally with AADT because high volume roads tend to receive more infrastructure investment. But when the outcome is crash severity rather than frequency, speed and road design tend to dominate. North Carolina's Vision Zero Action Plan (NCDOT, 2021) lists excessive speed as one of the top five causes of fatal crashes in the state, which is what this analysis tries to provide local evidence on.


# Data

## Datasets

This analysis uses three datasets:

**1. NHTSA Fatality Analysis Reporting System (FARS), 2018–2022**
FARS is a nationwide census of every motor vehicle crash resulting in at least one death within 30 days. Each record includes crash coordinates, road type, alcohol involvement, lighting conditions, fatality count, and urban or rural designation. I use all NC records (state FIPS code 37) from 2018 through 2022. Because FARS only includes fatal crashes, I cannot estimate fatality probability without also having data on crashes that were not fatal.

**2. NCDOT Statutory Speed Limits (GIS Shapefile)**
This dataset from NCDOT's Traffic Safety and GIS units contains the posted speed limit for every state maintained road segment in North Carolina as geographic line features. It has 108,566 segments with a `SpeedLimit` column in mph. It only covers state maintained roads, so some local roads are not included.

**3. NCDOT 2024 Annual Average Daily Traffic (AADT) Stations**
NCDOT maintains 48,687 AADT monitoring stations across state maintained highways, with traffic count data from 2002 to 2024. Many roads are only counted every other year, leaving gaps in individual year columns. To get a stable estimate, I averaged all available AADT values from 2016 through 2024 for each station.

## Data Preparation and Joining

All data preparation and analysis was done in R. FARS data was downloaded from the NHTSA FTP server. Speed limit and AADT shapefiles were loaded from local files. Crash points were projected to NC State Plane (EPSG:32119) and joined to the nearest speed limit segment and AADT station using `st_nearest_feature()` from the `sf` package. The full R code for this analysis is available at: https://github.com/TheLiamMacKinnon/FatalAccidentsReport

```{r setup}
# Load packages needed for this analysis
library(tidyverse)
library(sf)
library(tigris)
library(httr)
library(scales)
library(broom)
library(knitr)
library(kableExtra)

options(tigris_use_cache = TRUE)

# Set the project folder path - all shapefiles and downloaded data go here
project_dir <- "C:/Users/liamc/Traffic Report Liam MacKinnon"
dir.create(file.path(project_dir, "data"), showWarnings = FALSE, recursive = TRUE)
```

```{r download-fars}
# Download FARS crash data for NC from the NHTSA FTP server for each year 2018-2022
download_fars_year <- function(year, dir) {
  url  <- paste0("https://static.nhtsa.gov/nhtsa/downloads/FARS/",
                 year, "/National/FARS", year, "NationalCSV.zip")
  dest <- file.path(dir, "data", paste0("FARS_", year, ".zip"))
  out  <- file.path(dir, "data", paste0("FARS_", year))

  if (!file.exists(dest)) {
    GET(url, write_disk(dest, overwrite = TRUE), timeout(180))
  }

  files    <- unzip(dest, list = TRUE)$Name
  acc_file <- files[str_detect(tolower(files), "accident")]
  unzip(dest, files = acc_file, exdir = out, overwrite = FALSE)

  read_csv(file.path(out, acc_file), show_col_types = FALSE) %>%
    mutate(across(everything(), as.character)) %>%
    mutate(year = year)
}

# Bind all 5 years together - read as character first to avoid type conflicts across years
fars_raw <- map_dfr(2018:2022, download_fars_year, dir = project_dir) %>%
  mutate(
    STATE    = as.integer(STATE),
    FATALS   = as.integer(FATALS),
    LATITUDE = as.double(LATITUDE),
    LONGITUD = as.double(LONGITUD),
    LGT_COND = as.integer(LGT_COND),
    RUR_URB  = as.integer(RUR_URB),
    DRUNK_DR = as.integer(DRUNK_DR)
  ) %>%
  filter(STATE == 37)
```

```{r clean-fars}
# Clean up FARS - drop bad coordinates and recode the variables I need
fars <- fars_raw %>%
  filter(
    !is.na(LATITUDE), !is.na(LONGITUD),
    LATITUDE > 33, LATITUDE < 37,
    LONGITUD > -85, LONGITUD < -75
  ) %>%
  mutate(
    urban_rural = if_else(RUR_URB == 1, "Rural", "Urban"),
    alcohol     = if_else(DRUNK_DR >= 1, "Yes", "No")
  )
```

```{r load-speed-limits}
# Load the NCDOT speed limits shapefile and reproject to NC State Plane
speed_limits <- st_read(
  file.path(project_dir, "speed_limits", "NCDOT_Statutory_Speed_Limits.shp"),
  quiet = TRUE
) %>%
  st_transform(32119) %>%
  filter(!is.na(SpeedLimit), SpeedLimit > 0)
```

```{r load-aadt}
# Load NCDOT AADT stations and average across available years to handle biennial gaps
aadt <- st_read(
  file.path(project_dir, "aadt", "NCDOT_2024_AADT_Stations.shp"),
  quiet = TRUE
) %>%
  st_transform(32119) %>%
  mutate(across(starts_with("AADT_"), as.numeric)) %>%
  mutate(aadt_count = rowMeans(
    across(any_of(c("AADT_2016","AADT_2017","AADT_2018","AADT_2019",
                    "AADT_2020","AADT_2021","AADT_2022","AADT_2023","AADT_2024"))),
    na.rm = TRUE
  )) %>%
  filter(!is.na(aadt_count), !is.nan(aadt_count), aadt_count > 0)
```

```{r spatial-joins}
# Join each crash point to the nearest speed limit segment and nearest AADT station
fars_sf <- st_as_sf(fars, coords = c("LONGITUD", "LATITUDE"), crs = 4326) %>%
  st_transform(32119)

fars_speed <- st_join(
  fars_sf,
  speed_limits %>% select(SpeedLimit),
  join = st_nearest_feature
) %>%
  filter(!is.na(SpeedLimit))

fars_joined <- st_join(
  fars_speed,
  aadt %>% select(aadt_count),
  join = st_nearest_feature
) %>%
  filter(!is.na(aadt_count))

# Drop geometry for tables and regression, create speed category from joined SpeedLimit
crashes <- fars_joined %>%
  st_drop_geometry() %>%
  mutate(
    speed_cat = case_when(
      SpeedLimit <= 30                   ~ "<=30 mph",
      SpeedLimit > 30 & SpeedLimit <= 45 ~ "31-45 mph",
      SpeedLimit > 45 & SpeedLimit <= 55 ~ "46-55 mph",
      SpeedLimit > 55                    ~ "56+ mph",
      TRUE                               ~ NA_character_
    ) %>% factor(levels = c("<=30 mph", "31-45 mph", "46-55 mph", "56+ mph"))
  )
```

## Descriptive Statistics

```{r descriptive-stats}
# Summary statistics table for the full joined dataset
summary_tbl <- crashes %>%
  summarise(
    `Fatal Crashes`            = scales::comma(n()),
    `Total Fatalities`         = scales::comma(sum(FATALS, na.rm = TRUE)),
    `Mean Speed Limit (mph)`   = round(mean(SpeedLimit, na.rm = TRUE), 1),
    `Median Speed Limit (mph)` = median(SpeedLimit, na.rm = TRUE),
    `Mean AADT`                = scales::comma(round(mean(aadt_count, na.rm = TRUE))),
    `% Rural`                  = paste0(round(mean(urban_rural == "Rural", na.rm = TRUE) * 100, 1), "%"),
    `% Alcohol Involved`       = paste0(round(mean(alcohol == "Yes", na.rm = TRUE) * 100, 1), "%")
  ) %>%
  mutate(across(everything(), as.character)) %>%
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value")

kable(summary_tbl,
      caption = "Table 1: Descriptive Statistics, NC Fatal Crashes 2018 to 2022",
      align = c("l", "r"), booktabs = TRUE) %>%
  kable_styling(full_width = FALSE, latex_options = c("hold_position", "striped"))
```

```{r speed-summary-table}
# Break down fatal crashes and fatalities by speed limit category
speed_tbl <- crashes %>%
  filter(!is.na(speed_cat)) %>%
  group_by(`Speed Limit Category` = speed_cat) %>%
  summarise(
    `Fatal Crashes`            = scales::comma(n()),
    `Total Fatalities`         = scales::comma(sum(FATALS)),
    `Avg Fatalities per Crash` = round(mean(FATALS), 3),
    .groups = "drop"
  )

kable(speed_tbl,
      caption = "Table 2: Fatal Crashes and Fatalities by Speed Limit Category",
      align = c("l", "r", "r", "r"), booktabs = TRUE) %>%
  kable_styling(full_width = FALSE, latex_options = c("hold_position", "striped"))
```

The dataset includes `r scales::comma(nrow(crashes))` fatal crash records from North Carolina between 2018 and 2022. About `r round(mean(crashes$urban_rural == "Rural", na.rm=TRUE)*100, 1)`% of those crashes happened on rural roads, which lines up with what the national literature says about rural crash risk. Alcohol was a factor in roughly `r round(mean(crashes$alcohol == "Yes", na.rm=TRUE)*100, 1)`% of crashes.

# Methods

The analysis has three stages. First, I spatially joined each FARS crash point to the nearest speed limit road segment and nearest AADT monitoring station using `st_nearest_feature()` from the `sf` package. Second, I produced descriptive plots including a crash map, bar charts by speed category, and an urban or rural breakdown. Third, I ran an OLS regression with fatalities per crash as the outcome, using posted speed limit (mph) and log transformed AADT as the main predictors, controlling for urban or rural designation and alcohol involvement, and including an interaction term between speed limit and urban or rural setting.

# Results

## Spatial Distribution of Fatal Crashes

```{r map-crashes, fig.cap="Figure 1: Fatal crash locations in NC (2018-2022), colored by posted speed limit. High-speed crashes concentrate along rural highways and interstate corridors."}
# Map of all fatal crashes across NC, colored by the posted speed limit at each crash location
nc_counties <- counties(state = "NC", cb = TRUE, progress_bar = FALSE) %>%
  st_transform(32119)

ggplot() +
  geom_sf(data = nc_counties, fill = "gray93", color = "white", linewidth = 0.3) +
  geom_sf(data = fars_joined, aes(color = SpeedLimit),
          alpha = 0.45, size = 0.5, shape = 16) +
  scale_color_viridis_c(
    option = "plasma",
    name   = "Speed Limit\n(mph)",
    breaks = c(25, 35, 45, 55, 65)
  ) +
  labs(
    title    = "Fatal Crash Locations by Posted Speed Limit",
    subtitle = "North Carolina, 2018-2022",
    caption  = "Sources: NHTSA FARS; NCDOT Statutory Speed Limits"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(color = "gray40"),
    legend.position = "right"
  )
```

Figure 1 maps all fatal crashes across North Carolina, colored by the posted speed limit where the crash occurred. Crashes on higher-speed roads (55+ mph, shown in yellow) cluster along interstate corridors and rural US and NC highways, especially in the eastern coastal plain and the Piedmont. Lower-speed crashes (35 mph and below, shown in purple) are densely concentrated in the Charlotte, Raleigh-Durham, and Greensboro metro areas, which makes sense given the higher traffic density in those places.

## Fatalities by Speed Limit

```{r plot-speed-bar, fig.cap="Figure 2: Total traffic fatalities in NC by posted speed limit category, 2018-2022."}
# Bar chart showing total fatalities for each speed limit category
speed_sum <- crashes %>%
  filter(!is.na(speed_cat)) %>%
  group_by(speed_cat) %>%
  summarise(total_fats = sum(FATALS), n_crashes = n(), .groups = "drop")

ggplot(speed_sum, aes(x = speed_cat, y = total_fats, fill = speed_cat)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = scales::comma(total_fats)),
            vjust = -0.5, size = 3.8, fontface = "bold") +
  scale_fill_viridis_d(option = "plasma", begin = 0.15, end = 0.85) +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Total Traffic Fatalities by Speed Limit Category",
    subtitle = "North Carolina, 2018-2022",
    x        = "Posted Speed Limit",
    y        = "Total Fatalities",
    caption  = "Sources: NHTSA FARS; NCDOT Statutory Speed Limits"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold"),
    plot.subtitle      = element_text(color = "gray40"),
    panel.grid.major.x = element_blank()
  )
```

Figure 2 shows that the 46–55 mph and 31–45 mph categories account for the most total fatalities. That said, raw totals are not the whole picture. Roads with higher speed limits also tend to carry more traffic, so some of what we see here is just greater exposure rather than higher crash severity per crash. The regression below tries to separate those effects.

## Urban vs. Rural Breakdown

```{r plot-urban-rural, fig.cap="Figure 3: Fatal crash counts by speed limit category and urban or rural setting."}
# Grouped bar chart comparing rural and urban fatal crashes across speed limit categories
crashes %>%
  filter(!is.na(speed_cat), !is.na(urban_rural)) %>%
  ggplot(aes(x = speed_cat, fill = urban_rural)) +
  geom_bar(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Rural" = "#d62728", "Urban" = "#3182bd")) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Fatal Crashes by Speed Limit and Urban/Rural Setting",
    subtitle = "North Carolina, 2018-2022",
    x        = "Posted Speed Limit",
    y        = "Number of Fatal Crashes",
    fill     = NULL,
    caption  = "Sources: NHTSA FARS; NCDOT Statutory Speed Limits"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold"),
    plot.subtitle      = element_text(color = "gray40"),
    panel.grid.major.x = element_blank(),
    legend.position    = "top"
  )
```

Figure 3 makes it pretty clear that rural crashes dominate the higher speed limit categories. The 46–55 mph and 56+ mph bars are almost entirely rural, while urban crashes are concentrated in the lower speed limit zones. This is what you'd expect given that rural roads tend to have higher posted speed limits, and rural crashes are more likely to be fatal since emergency response takes longer and the roads are less forgiving.

## Regression Results

```{r regression}
# OLS regression of fatalities per crash on speed limit, AADT, urban or rural, and alcohol
crashes <- crashes %>%
  mutate(
    urban_rural = factor(urban_rural),
    alcohol     = factor(alcohol)
  )

model <- lm(
  FATALS ~ SpeedLimit + log(aadt_count + 1) + urban_rural + alcohol +
           SpeedLimit:urban_rural,
  data = crashes
)

model_tidy <- tidy(model, conf.int = TRUE) %>%
  mutate(
    term = str_replace_all(term, c(
      "SpeedLimit"                  = "Speed Limit (mph)",
      "log\\(aadt_count \\+ 1\\)"  = "Log(AADT)",
      "urban_ruralUrban"            = "Urban (ref: Rural)",
      "alcoholYes"                  = "Alcohol Involved",
      "SpeedLimit:urban_ruralUrban" = "Speed Limit x Urban"
    )),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      p.value < 0.1   ~ ".",
      TRUE            ~ ""
    )
  )

model_fit <- glance(model)

model_tidy %>%
  select(Term = term, Estimate = estimate, `Std. Error` = std.error,
         `t-stat` = statistic, `p-value` = p.value, `Sig.` = sig) %>%
  mutate(across(where(is.numeric), ~round(.x, 4))) %>%
  kable(caption = "Table 3: OLS Regression, Predictors of Fatal Crash Severity",
        align = c("l", "r", "r", "r", "r", "c"), booktabs = TRUE) %>%
  kable_styling(full_width = FALSE, latex_options = c("hold_position", "striped")) %>%
  footnote(general = paste0(
    "*** p<0.001  ** p<0.01  * p<0.05  . p<0.1 | ",
    "Ref: Rural, No alcohol | ",
    "R2 = ", round(model_fit$r.squared, 3),
    " | n = ", scales::comma(model_fit$nobs)
  ), escape = FALSE)
```

```{r plot-coef, fig.cap="Figure 4: Regression coefficients with 95% confidence intervals. Red = positive association with fatalities; blue = negative."}
# Coefficient plot to visualize the direction and uncertainty of each predictor
model_tidy %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = estimate, y = reorder(term, estimate),
             xmin = conf.low, xmax = conf.high, color = estimate > 0)) +
  geom_pointrange(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(values = c("TRUE" = "#d62728", "FALSE" = "#3182bd"), guide = "none") +
  labs(
    title    = "Regression Coefficients: Predictors of Crash Fatalities",
    subtitle = paste0("OLS with 95% Confidence Intervals  |  R² = ",
                      round(model_fit$r.squared, 3),
                      "  |  n = ", scales::comma(model_fit$nobs)),
    x = "Estimated Effect on Fatalities per Crash",
    y = NULL,
    caption = "Sources: NHTSA FARS; NCDOT Speed Limits & AADT"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(color = "gray40", size = 10))
```

Table 3 and Figure 4 show the regression results. To put the numbers in plain terms, I'm comparing them to the mean of about 1.086 fatalities per crash across all crashes in the dataset.

**Alcohol involvement** is the strongest predictor and the only one that's statistically significant. Crashes involving a drunk driver result in about 2.9% more fatalities per crash compared to crashes without alcohol. That's a meaningful difference and the clearest policy signal in the data.

**Posted speed limit** shows a positive but small association. A 10 mph increase in the posted speed limit is associated with roughly 0.7% more fatalities per crash, after controlling for traffic volume and other factors. The direction is consistent with what the Power Model predicts, but the effect is pretty modest.

**Traffic volume (AADT)** also has a positive association. Moving from a low volume road to a high volume road is associated with roughly 1.5% more fatalities per crash, which might reflect the more complex, multi-vehicle collisions that happen on busier roads.

**Urban roads** average about 3.6% fewer fatalities per crash than rural roads, which makes sense given faster emergency response times and generally lower speed limits in cities. The interaction between speed limit and urban or rural setting doesn't show a meaningful difference, suggesting speed limits affect fatality risk similarly in both settings.

The R² of `r round(model_fit$r.squared, 3)` means the model only explains about `r round(model_fit$r.squared*100, 1)`% of the variation in fatalities per crash. That is low, but expected. Individual crash outcomes depend on a lot of things I do not have data on, like whether people were wearing seatbelts, what kind of vehicles were involved, and exactly how the crash happened. The low R² does not mean the directional findings are wrong, just that there is a lot more going on that this model does not capture.

# Policy Implications

The results point to three clear priorities for NCDOT and the NC Governor's Highway Safety Program under Vision Zero.

First, **impaired driving enforcement** is the most evidence backed intervention. Alcohol is the only statistically significant predictor in the model, associated with 2.9% more fatalities per crash. Sobriety checkpoints and DWI enforcement on rural highways are the clearest levers available.

Second, **speed limit review on rural corridors** is warranted. A 10 mph reduction in posted speed limit is associated with roughly 0.7% fewer fatalities per crash. That's small per crash, but meaningful in aggregate across thousands of crashes annually. Rural highways with 55+ mph limits and long EMS response times should be the priority.

Third, **infrastructure improvements on high volume roads** could reduce severity where traffic exposure is highest. Median barriers, rumble strips, and better lighting on busy corridors would help.

One key limitation is that this analysis only uses fatal crashes. Without data on all crashes, I cannot compute true fatality rates per vehicle mile traveled. Incorporating the full NC crash database would make future analysis much stronger.

# Conclusion

This report looked at the relationship between posted speed limits, traffic volume, and fatal crash severity across North Carolina from 2018 through 2022. Crashes on high speed roads cluster along rural corridors in the eastern coastal plain and Piedmont, while lower-speed crashes concentrate in the major metro areas. The regression shows alcohol involvement is the strongest predictor at 2.9% more fatalities per crash, followed by traffic volume at about 1.5% and speed limit at about 0.7% per 10 mph. Urban roads average 3.6% fewer fatalities than rural. For Vision Zero, impaired driving enforcement is the clearest priority, with rural speed limit reductions and infrastructure on high volume roads as supporting interventions.

# Bibliography

- Elvik, R., Christensen, P., & Amundsen, A. (2004). *Speed and road accidents: An evaluation of the Power Model*. TØI Report 740/2004. Institute of Transport Economics.
- Farmer, C. M. (2019). Relationship of traffic fatality rates to maximum state speed limits. *Traffic Injury Prevention, 18*(4), 375–380.
- Lord, D., & Mannering, F. (2010). The statistical analysis of crash-frequency data: A review and assessment of methodological alternatives. *Transportation Research Part A, 44*(5), 291–305.
- NCDOT. (2021). *North Carolina Vision Zero Action Plan*. NC Department of Transportation.
- NCDOT. (2023). *North Carolina Crash Facts 2022*. NC Department of Transportation, Division of Motor Vehicles.
- Nilsson, G. (2004). *Traffic safety dimensions and the Power Model to describe the effect of speed on safety*. Bulletin 221. Lund Institute of Technology.
- Zwerling, C., Peek-Asa, C., Whitten, P. S., Choi, S. W., Sprince, N. L., & Jones, M. P. (2005). Fatal motor vehicle crashes in rural and urban areas. *Injury Prevention, 11*(3), 127–132.

**Data Sources**

- National Highway Traffic Safety Administration (NHTSA). (2023). *Fatality Analysis Reporting System (FARS), 2018–2022* [Dataset]. U.S. Department of Transportation. https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars
- North Carolina Department of Transportation (NCDOT), Traffic Safety Unit & GIS Unit. (2023). *NCDOT Statutory Speed Limits* [GIS Shapefile]. https://opendata.rcmrd.org/datasets/NCDOT::ncdot-statutory-speed-limits
- North Carolina Department of Transportation (NCDOT), Traffic Survey Group. (2025). *NCDOT 2024 Annual Average Daily Traffic (AADT) Stations* [GIS Shapefile]. https://connect.ncdot.gov/resources/State-Mapping/Pages/Traffic-Volume-Maps.aspx

# AI Usage

| Usage | Tool Used | How Edited |
|---|---|---|
| Code | Claude | A lot of formatting and syntax, making my code more uniform |
| Brainstorming and Idea Generation | Claude | I came up with the original idea but found my results lackluster and Claude helped me incorporate Alcohol into the paper, it was pretty easy because FARS already included alcohol related fatal accidents as a statistic in my data set. |
| Research | Claude |  Helped with finding sources to tie into my research |
