---
title: "Speed Limits, Traffic Volume, and Fatal Crash Risk in North Carolina"
subtitle: "PLAN 372 Final Project"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    fig-cap-location: bottom
    number-sections: true
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
    fontsize: 12pt
    linestretch: 2
    latex-engine: xelatex
    keep-tex: false
execute:
  warning: false
  message: false
  echo: false
  cache: true
---

# Introduction and Motivation

Traffic fatalities represent one of the most preventable causes of death in the United States. In North Carolina, hundreds of people are killed in motor vehicle crashes every year, placing enormous burdens on public health systems, emergency services, and families across the state. Unlike many public health crises, traffic deaths are largely predictable and preventable — they follow patterns tied to road design, driver behavior, and the physical environment in which crashes occur.

Two factors consistently appear in the traffic safety literature as central determinants of crash severity: **posted speed limits** and **traffic volume**. Higher speed limits reduce reaction time and dramatically increase the kinetic energy transferred during a collision, making fatal outcomes far more likely. Traffic volume, measured as Annual Average Daily Traffic (AADT), determines crash *exposure* — roads with more vehicles see more crashes simply because more interactions occur. Understanding how these two factors jointly contribute to fatality risk is essential for designing effective road safety policy.

This report investigates the relationship between posted speed limits, traffic volume, and traffic fatality rates across North Carolina using data from 2018–2022. The central research question is: **Are posted speed limits and traffic volume associated with fatal crash outcomes in North Carolina, and does this relationship differ between urban and rural settings?**

The findings have direct implications for North Carolina's Vision Zero initiative, which aims to eliminate all traffic fatalities on state roads. If speed limits are a significant predictor of fatality risk after controlling for traffic volume and other factors, this would support targeted speed limit reductions on high-risk corridors. If traffic volume is the stronger driver, the policy focus should shift toward infrastructure improvements and congestion management on high-volume roads.

# Literature Review

The relationship between vehicle speed and crash severity is one of the most well-established findings in traffic safety research. Nilsson (2004) developed the "Power Model," which estimates that a 1% increase in mean traffic speed results in approximately a 2% increase in injury crashes and a 4% increase in fatal crashes. This nonlinear relationship — where small speed increases produce disproportionately large fatality increases — underpins most modern speed management policy. The model has been validated across multiple countries and road types (Elvik et al., 2004).

In the United States context, Farmer (2019) found that raising speed limits on U.S. interstates and freeways between 1993 and 2017 was associated with an 8.5% increase in fatality rates on those roads, even as fatality rates fell on roads where limits were unchanged. This natural experiment provides compelling evidence that posted speed limits — not just actual travel speeds — directly affect fatality outcomes.

Rural roads present a particular concern in North Carolina. Research by Zwerling et al. (2005) found that rural residents in the U.S. face significantly higher traffic fatality rates than urban residents, due to a combination of higher posted speeds, longer emergency response times, narrower roadways, and fewer safety features such as median barriers and rumble strips. North Carolina's rural counties, particularly in the eastern coastal plain and western mountains, exhibit some of the highest crash fatality rates in the state (NCDOT, 2023).

The role of traffic volume in crash frequency follows a more complex pattern. Exposure models in the transportation safety literature, such as those reviewed by Lord and Mannering (2010), generally find that crashes scale sub-linearly with AADT — doubling traffic volume less than doubles crash counts — because higher-volume roads tend to attract greater infrastructure investment and enforcement. However, when crash *severity* rather than frequency is the outcome, the relationship with volume becomes less clear, and speed and road design factors tend to dominate.

North Carolina's Vision Zero Action Plan (NCDOT, 2021) explicitly identifies excessive speed as one of the five primary contributing factors to fatal crashes in the state, alongside impaired driving, failure to wear seatbelts, distracted driving, and aggressive driving. The plan calls for data-driven identification of high-risk corridors and evidence-based speed limit setting as core strategies, making the analysis presented here directly relevant to ongoing policy work.

**Bibliography**

- Elvik, R., Christensen, P., & Amundsen, A. (2004). *Speed and road accidents: An evaluation of the Power Model*. TØI Report 740/2004. Institute of Transport Economics.
- Farmer, C. M. (2019). Relationship of traffic fatality rates to maximum state speed limits. *Traffic Injury Prevention, 18*(4), 375–380.
- Lord, D., & Mannering, F. (2010). The statistical analysis of crash-frequency data: A review and assessment of methodological alternatives. *Transportation Research Part A, 44*(5), 291–305.
- NCDOT. (2021). *North Carolina Vision Zero Action Plan*. NC Department of Transportation.
- NCDOT. (2023). *North Carolina Crash Facts 2022*. NC Department of Transportation, Division of Motor Vehicles.
- Nilsson, G. (2004). *Traffic safety dimensions and the Power Model to describe the effect of speed on safety*. Bulletin 221. Lund Institute of Technology.
- Zwerling, C., Peek-Asa, C., Whitten, P. S., Choi, S. W., Sprince, N. L., & Jones, M. P. (2005). Fatal motor vehicle crashes in rural and urban areas. *Injury Prevention, 11*(3), 127–132.

# Data

## Datasets

This analysis draws on three datasets:

**1. NHTSA Fatality Analysis Reporting System (FARS), 2018–2022**
FARS is a nationwide census of all motor vehicle crashes resulting in at least one fatality within 30 days of the incident. Maintained by the National Highway Traffic Safety Administration, FARS has been collected annually since 1975. Each crash record includes precise geographic coordinates (latitude/longitude), road classification, weather and lighting conditions, alcohol involvement, number of fatalities, and urban/rural designation. This analysis uses all NC records (state FIPS code 37) for 2018–2022. Because FARS covers only fatal crashes, the analysis cannot directly estimate fatality probability per crash without a full crash database as the denominator.

**2. NCDOT Statutory Speed Limits (GIS Shapefile)**
Published by NCDOT's Traffic Safety and GIS Units, this dataset provides posted speed limits for all state-maintained road segments in North Carolina as geographic line features, derived from official traffic ordinances. It contains 108,566 road segments with a key column `SpeedLimit` (mph). Coverage is limited to state-maintained roads; locally maintained roads may not be fully represented.

**3. NCDOT 2024 Annual Average Daily Traffic (AADT) Stations**
The NCDOT Traffic Survey Group publishes AADT monitoring station data covering 48,687 stations across all state-maintained highways in North Carolina, with traffic count history from 2002 to 2024. The dataset includes yearly AADT columns (`AADT_2002` through `AADT_2024`). Because AADT stations follow a biennial collection cycle for many road types, individual year columns frequently contain NaN values for years a station was not counted. To address this, the analysis averages all available AADT values from 2018–2022 to produce a stable, crash-period-matched traffic volume estimate for each station. Because AADT stations are point features, some crash locations may be matched to a nearby station on a different road, introducing minor measurement error in the traffic volume variable.

## Data Preparation and Joining

All data preparation, spatial joining, and analysis was performed in R. The FARS data was downloaded directly from the NHTSA FTP server. Speed limits and AADT were loaded from NCDOT shapefiles. Crash points were converted to spatial features, projected to NC State Plane (EPSG:32119), and joined to the nearest speed limit road segment and nearest AADT station using `st_nearest_feature()`. A 50-meter distance threshold filtered out crashes that could not be reliably matched to a state-maintained road.

```{r setup}
# Load all required packages
library(tidyverse)    # data manipulation and ggplot2
library(sf)           # spatial data handling
library(tigris)       # NC county shapefiles from Census
library(httr)         # HTTP requests for downloading FARS files
library(viridis)      # colorblind-friendly color palettes
library(scales)       # axis label formatting helpers
library(broom)        # tidy regression output
library(knitr)        # table rendering
library(kableExtra)   # enhanced table styling

options(tigris_use_cache = TRUE)

# -----------------------------------------------------------------------
# SET YOUR PROJECT FOLDER PATH HERE
# All shapefiles and downloaded FARS data will be read from/saved here
# -----------------------------------------------------------------------
project_dir <- "C:/Users/liamc/PLAN372_FinalReport"

# Create subdirectories if they don't exist
dir.create(file.path(project_dir, "data"), showWarnings = FALSE, recursive = TRUE)
```

```{r download-fars}
# -----------------------------------------------------------------------
# Download FARS data for NC (2018-2022) directly from NHTSA
# Each year is a zip file; we extract the accident-level CSV and filter to NC
# Files are cached locally so this only downloads once
# -----------------------------------------------------------------------

download_fars_year <- function(year, dir) {
  url  <- paste0("https://static.nhtsa.gov/nhtsa/downloads/FARS/",
                 year, "/National/FARS", year, "NationalCSV.zip")
  dest <- file.path(dir, "data", paste0("FARS_", year, ".zip"))
  out  <- file.path(dir, "data", paste0("FARS_", year))

  # Only download if not already cached
  if (!file.exists(dest)) {
    message("Downloading FARS ", year, "...")
    GET(url, write_disk(dest, overwrite = TRUE), timeout(180))
  }

  # Find and extract the accident-level CSV from the zip archive
  files    <- unzip(dest, list = TRUE)$Name
  acc_file <- files[str_detect(tolower(files), "accident")]
  unzip(dest, files = acc_file, exdir = out, overwrite = FALSE)

  read_csv(file.path(out, acc_file), show_col_types = FALSE) %>%
    mutate(across(everything(), as.character)) %>%  # coerce all to character first to avoid type conflicts across years
    mutate(year = year)
}

# Download all 5 years — bind as character first to handle year-to-year type
# inconsistencies (e.g. LATITUDENAME is double in some years, character in others),
# then convert key numeric columns back after binding
fars_raw <- map_dfr(2018:2022, download_fars_year, dir = project_dir) %>%
  mutate(
    STATE    = as.integer(STATE),
    FATALS   = as.integer(FATALS),
    LATITUDE = as.double(LATITUDE),
    LONGITUD = as.double(LONGITUD),
    LGT_COND = as.integer(LGT_COND),
    RUR_URB  = as.integer(RUR_URB),
    DRUNK_DR = as.integer(DRUNK_DR)
  ) %>%
  filter(STATE == 37)

message("Raw NC FARS records loaded: ", nrow(fars_raw))
```

```{r clean-fars}
# -----------------------------------------------------------------------
# Clean FARS: drop invalid coordinates, recode categorical variables
# -----------------------------------------------------------------------

fars <- fars_raw %>%
  filter(
    !is.na(LATITUDE), !is.na(LONGITUD),
    LATITUDE  >  33,  LATITUDE  <  37,   # valid NC latitude bounds
    LONGITUD  > -85,  LONGITUD  < -75    # valid NC longitude bounds
  ) %>%
  mutate(
    # Recode numeric lighting code to readable label
    lighting = case_when(
      LGT_COND == 1 ~ "Daylight",
      LGT_COND == 2 ~ "Dark - Not Lighted",
      LGT_COND == 3 ~ "Dark - Lighted",
      LGT_COND == 4 ~ "Dawn/Dusk",
      TRUE          ~ "Unknown"
    ),
    # Simplify weather into broad categories
    weather = case_when(
      WEATHERNAME %in% c("Clear", "Cloudy")     ~ WEATHERNAME,
      str_detect(WEATHERNAME, "Rain")            ~ "Rain",
      str_detect(WEATHERNAME, "Snow|Sleet|Ice") ~ "Snow/Ice",
      TRUE                                        ~ "Other/Unknown"
    ),
    # Urban vs rural from FARS RUR_URB field (1 = rural, 2 = urban)
    urban_rural = if_else(RUR_URB == 1, "Rural", "Urban"),
    # Alcohol involvement: any drunk driver = Yes
    alcohol = if_else(DRUNK_DR >= 1, "Yes", "No"),
    # speed_cat is created after the spatial join using the road-level
    # SpeedLimit column from NCDOT (more accurate than the FARS vehicle field)
  )

message("Clean FARS records: ", nrow(fars))
```

```{r load-speed-limits}
# -----------------------------------------------------------------------
# Load NCDOT Statutory Speed Limits shapefile
# Confirmed columns: SpeedLimit (mph), RouteClass, MaintCntyC, geometry
# CRS: EPSG:2264 (NC State Plane feet) -> reprojected to EPSG:32119 (meters)
#
# Place all shapefile components (.shp, .dbf, .shx, .prj, .cpg) in:
#   C:/Users/liamc/PLAN372_FinalReport/speed_limits/
# -----------------------------------------------------------------------

speed_limits <- st_read(
  file.path(project_dir, "speed_limits", "NCDOT_Statutory_Speed_Limits.shp"),
  quiet = TRUE
) %>%
  st_transform(32119) %>%                        # reproject to NC State Plane (meters)
  filter(!is.na(SpeedLimit), SpeedLimit > 0)     # drop missing/zero speed limits

message("Speed limit segments loaded: ", nrow(speed_limits))
```

```{r load-aadt}
# -----------------------------------------------------------------------
# Load NCDOT 2024 AADT Stations shapefile
# Confirmed columns: LocationID, County, Route, AADT_2018 through AADT_2022,
#                    Latitude, Longitude, RTE_CLS, FUNC_CLS, geometry
# CRS: EPSG:4326 (WGS84) -> reprojected to EPSG:32119
#
# Note: Many stations use a biennial collection cycle, so individual year
# columns often contain NaN. We average all available years from 2018-2022
# to create a stable AADT estimate that matches our FARS study period.
#
# Place all shapefile components in:
#   C:/Users/liamc/PLAN372_FinalReport/aadt/
# -----------------------------------------------------------------------

aadt <- st_read(
  file.path(project_dir, "aadt", "NCDOT_2024_AADT_Stations.shp"),
  quiet = TRUE
) %>%
  st_transform(32119) %>%
  # AADT columns are stored as character in the shapefile — convert to numeric first
  mutate(across(c(AADT_2018, AADT_2019, AADT_2020, AADT_2021, AADT_2022),
                as.numeric)) %>%
  # Average across 2018-2022, ignoring NA years due to biennial collection cycle
  mutate(aadt_count = rowMeans(
    across(c(AADT_2018, AADT_2019, AADT_2020, AADT_2021, AADT_2022)),
    na.rm = TRUE
  )) %>%
  filter(!is.na(aadt_count), !is.nan(aadt_count), aadt_count > 0)

message("AADT stations loaded: ", nrow(aadt))
```

```{r spatial-joins}
# -----------------------------------------------------------------------
# Spatial Join 1: FARS crash points -> nearest NCDOT speed limit road segment
#   Appends: SpeedLimit (posted speed in mph)
#
# Spatial Join 2: Result -> nearest NCDOT AADT monitoring station
#   Appends: aadt_count (avg daily vehicles 2018-2022)
#
# Quality filter: drop crashes matched to a road segment > 50m away,
#   which indicates GPS imprecision or a locally maintained road not in NCDOT
# -----------------------------------------------------------------------

# Convert FARS to sf point layer and reproject to NC State Plane
fars_sf <- st_as_sf(fars,
                    coords = c("LONGITUD", "LATITUDE"),
                    crs    = 4326) %>%
  st_transform(32119)

# --- Join 1: speed limits ---
message("Joining crashes to speed limits...")
fars_speed <- st_join(
  fars_sf,
  speed_limits %>% select(SpeedLimit),
  join = st_nearest_feature
)

# Calculate distance to nearest road segment and apply 50m quality filter
nearest_idx       <- st_nearest_feature(fars_sf, speed_limits)
nearest_dist      <- st_distance(fars_sf, speed_limits[nearest_idx, ],
                                 by_element = TRUE)
fars_speed$dist_m <- as.numeric(nearest_dist)
fars_speed        <- fars_speed %>% filter(dist_m <= 50)

message("After 50m filter: ", nrow(fars_speed), " crashes retained")

# --- Join 2: AADT ---
message("Joining crashes to AADT stations...")
fars_joined <- st_join(
  fars_speed,
  aadt %>% select(aadt_count, FUNC_CLS, RTE_CLS),
  join = st_nearest_feature
) %>%
  filter(!is.na(aadt_count))

message("Final joined dataset: ", nrow(fars_joined), " crashes")

# Drop geometry for regression and table work (keep fars_joined sf for mapping)
# Create speed_cat here using SpeedLimit from the NCDOT spatial join
# (more reliable than FARS vehicle-level speed field which varies by year)
crashes <- fars_joined %>%
  st_drop_geometry() %>%
  mutate(
    speed_cat = case_when(
      SpeedLimit <= 30                            ~ "<=30 mph",
      SpeedLimit > 30 & SpeedLimit <= 45          ~ "31-45 mph",
      SpeedLimit > 45 & SpeedLimit <= 55          ~ "46-55 mph",
      SpeedLimit > 55                             ~ "56+ mph",
      TRUE                                         ~ NA_character_
    ) %>% factor(levels = c("<=30 mph", "31-45 mph", "46-55 mph", "56+ mph"))
  )
```

## Descriptive Statistics

```{r descriptive-stats}
# Overall summary statistics for the joined crash dataset
summary_tbl <- crashes %>%
  summarise(
    `Fatal Crashes`            = scales::comma(n()),
    `Total Fatalities`         = scales::comma(sum(FATALS, na.rm = TRUE)),
    `Mean Speed Limit (mph)`   = round(mean(SpeedLimit, na.rm = TRUE), 1),
    `Median Speed Limit (mph)` = median(SpeedLimit, na.rm = TRUE),
    `Mean AADT`                = scales::comma(round(mean(aadt_count, na.rm = TRUE))),
    `% Rural`                  = paste0(round(mean(urban_rural == "Rural",
                                                    na.rm = TRUE) * 100, 1), "%"),
    `% Alcohol Involved`       = paste0(round(mean(alcohol == "Yes",
                                                    na.rm = TRUE) * 100, 1), "%")
  ) %>%
  mutate(across(everything(), as.character)) %>%  # coerce all to character for pivot
  pivot_longer(everything(), names_to = "Statistic", values_to = "Value")

kable(summary_tbl,
      caption = "Table 1: Descriptive Statistics — NC Fatal Crashes, 2018–2022",
      align   = c("l", "r"),
      booktabs = TRUE) %>%
  kable_styling(full_width = FALSE,
                latex_options = c("hold_position", "striped"))
```

```{r speed-summary-table}
# Fatality counts broken out by posted speed limit category
speed_tbl <- crashes %>%
  filter(!is.na(speed_cat)) %>%
  group_by(`Speed Limit Category` = speed_cat) %>%
  summarise(
    `Fatal Crashes`            = scales::comma(n()),
    `Total Fatalities`         = scales::comma(sum(FATALS)),
    `Avg Fatalities per Crash` = round(mean(FATALS), 3),
    .groups = "drop"
  )

kable(speed_tbl,
      caption = "Table 2: Fatal Crashes and Fatalities by Speed Limit Category",
      align   = c("l", "r", "r", "r"),
      booktabs = TRUE) %>%
  kable_styling(full_width = FALSE,
                latex_options = c("hold_position", "striped"))
```

The dataset contains `r scales::comma(nrow(crashes))` fatal crash records across North Carolina from 2018–2022 after spatial filtering. The majority of fatal crashes (`r round(mean(crashes$urban_rural == "Rural", na.rm=TRUE)*100, 1)`%) occurred on rural roads, consistent with national patterns. Alcohol was involved in approximately `r round(mean(crashes$alcohol == "Yes", na.rm=TRUE)*100, 1)`% of fatal crashes.

# Methods

The analysis proceeds in three stages: spatial data integration, descriptive visualization, and regression modeling.

**Spatial joining** was performed using the `sf` package in R. FARS crash coordinates were converted to spatial point features and projected to NC State Plane (EPSG:32119) to match the NCDOT datasets. Each crash was joined to the nearest speed limit road segment and nearest AADT monitoring station using `st_nearest_feature()`. A 50-meter maximum distance threshold excluded records that could not be reliably matched to a state-maintained road. Because NCDOT AADT stations use a biennial collection cycle, individual year columns frequently contain missing values; the analysis addresses this by averaging all available AADT values from 2018–2022 for each station, producing a stable traffic volume estimate matched to the FARS study period.

**Descriptive analysis** includes summary statistics, bar charts of fatalities by speed limit category, and point maps of crash locations across NC colored by speed limit.

**Regression analysis** uses ordinary least squares (OLS) with total fatalities per crash as the outcome. The primary explanatory variables are posted speed limit (continuous, mph) and log-transformed average AADT (to account for the right-skewed distribution of traffic counts). Control variables include urban/rural designation, alcohol involvement, and lighting conditions. An interaction term between speed limit and urban/rural designation tests whether the speed-fatality relationship is stronger on rural roads.

# Results

## Spatial Distribution of Fatal Crashes

```{r map-crashes, fig.cap="Figure 1: Fatal crash locations in NC (2018-2022), colored by posted speed limit. High-speed crashes concentrate along rural highways and interstate corridors."}
# Download NC county boundaries from Census TIGER for basemap
nc_counties <- counties(state = "NC", cb = TRUE, progress_bar = FALSE) %>%
  st_transform(32119)

ggplot() +
  geom_sf(data = nc_counties,
          fill = "gray93", color = "white", linewidth = 0.3) +
  geom_sf(data  = fars_joined,
          aes(color = SpeedLimit),
          alpha = 0.45, size = 0.5, shape = 16) +
  scale_color_viridis_c(
    option = "plasma",
    name   = "Speed Limit\n(mph)",
    breaks = c(25, 35, 45, 55, 65)
  ) +
  labs(
    title    = "Fatal Crash Locations by Posted Speed Limit",
    subtitle = "North Carolina, 2018-2022",
    caption  = "Sources: NHTSA FARS; NCDOT Statutory Speed Limits"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    plot.subtitle   = element_text(color = "gray40"),
    legend.position = "right"
  )
```

Figure 1 shows the spatial distribution of all fatal crashes, colored by posted speed limit. Crashes on higher-speed roads (55+ mph, shown in yellow) concentrate along interstate corridors and rural US/NC highways, particularly in the eastern coastal plain and Piedmont. Lower-speed crashes (35 mph and below, shown in purple) cluster densely in the Charlotte, Raleigh-Durham, and Greensboro metro areas, reflecting higher urban traffic density.

## Fatalities by Speed Limit

```{r plot-speed-bar, fig.cap="Figure 2: Total traffic fatalities in NC by posted speed limit category, 2018-2022."}
speed_sum <- crashes %>%
  filter(!is.na(speed_cat)) %>%
  group_by(speed_cat) %>%
  summarise(total_fats = sum(FATALS), n_crashes = n(), .groups = "drop")

ggplot(speed_sum, aes(x = speed_cat, y = total_fats, fill = speed_cat)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = scales::comma(total_fats)),
            vjust = -0.5, size = 3.8, fontface = "bold") +
  scale_fill_viridis_d(option = "plasma", begin = 0.15, end = 0.85) +
  scale_y_continuous(labels = comma,
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Total Traffic Fatalities by Speed Limit Category",
    subtitle = "North Carolina, 2018-2022",
    x        = "Posted Speed Limit",
    y        = "Total Fatalities",
    caption  = "Sources: NHTSA FARS; NCDOT Statutory Speed Limits"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold"),
    plot.subtitle      = element_text(color = "gray40"),
    panel.grid.major.x = element_blank()
  )
```

Figure 2 shows that the largest number of total fatalities occurred on 46–55 mph and 56+ mph roads. However, these totals conflate crash frequency and severity — higher-speed roads may generate more crashes simply due to greater exposure. The regression below disentangles these effects.

## Speed Limit and AADT vs. Fatalities

```{r plot-scatter-speed, fig.cap="Figure 3: Relationship between posted speed limit and fatalities per crash, with OLS regression line and 95% confidence interval."}
ggplot(crashes, aes(x = SpeedLimit, y = FATALS)) +
  geom_jitter(alpha = 0.12, width = 1.2, color = "#3182bd", size = 0.7) +
  geom_smooth(method = "lm", color = "#d62728", se = TRUE, linewidth = 1.1) +
  labs(
    title    = "Posted Speed Limit vs. Crash Fatalities",
    subtitle = "North Carolina Fatal Crashes, 2018-2022",
    x        = "Posted Speed Limit (mph)",
    y        = "Fatalities per Crash",
    caption  = "Sources: NHTSA FARS; NCDOT Statutory Speed Limits"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "gray40"),
        plot.title.position = "plot")
```

```{r plot-scatter-aadt, fig.cap="Figure 4: Relationship between average traffic volume (AADT, log scale) and fatalities per crash."}
ggplot(crashes, aes(x = aadt_count, y = FATALS)) +
  geom_jitter(alpha = 0.12, color = "#31a354", size = 0.7) +
  geom_smooth(method = "lm", color = "#d62728", se = TRUE, linewidth = 1.1) +
  scale_x_log10(labels = comma) +
  labs(
    title    = "Traffic Volume (AADT) vs. Crash Fatalities",
    subtitle = "North Carolina Fatal Crashes, 2018-2022",
    x        = "Average Annual Daily Traffic 2018-2022 (log scale)",
    y        = "Fatalities per Crash",
    caption  = "Sources: NHTSA FARS; NCDOT 2024 AADT Stations"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "gray40"),
        plot.title.position = "plot")
```

## Urban vs. Rural Breakdown

```{r plot-urban-rural, fig.cap="Figure 5: Fatal crash counts by speed limit category and urban/rural setting. Rural crashes dominate higher speed limit categories."}
crashes %>%
  filter(!is.na(speed_cat), !is.na(urban_rural)) %>%
  ggplot(aes(x = speed_cat, fill = urban_rural)) +
  geom_bar(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Rural" = "#d62728", "Urban" = "#3182bd")) +
  scale_y_continuous(labels = comma) +
  labs(
    title    = "Fatal Crashes by Speed Limit and Urban/Rural Setting",
    subtitle = "North Carolina, 2018-2022",
    x        = "Posted Speed Limit",
    y        = "Number of Fatal Crashes",
    fill     = NULL,
    caption  = "Sources: NHTSA FARS; NCDOT Statutory Speed Limits"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold"),
    plot.subtitle      = element_text(color = "gray40"),
    panel.grid.major.x = element_blank(),
    legend.position    = "top"
  )
```

Figure 5 reveals that rural crashes dominate the 46–55 mph and 56+ mph categories, while urban crashes concentrate in lower speed limit zones. This is consistent with expectations that rural roads carry higher posted speed limits and that rural crashes are more likely to be fatal due to longer emergency response times and less forgiving road geometry.

## Regression Results

```{r regression}
# OLS regression: fatalities per crash ~ speed limit + log(AADT) + controls
# Interaction term: tests whether the speed-fatality relationship is
# stronger on rural roads compared to urban roads
model <- lm(
  FATALS ~ SpeedLimit +
           log(aadt_count + 1) +
           urban_rural +
           alcohol +
           lighting +
           SpeedLimit:urban_rural,   # interaction term
  data = crashes
)

# Relabel coefficient names for clean display
model_tidy <- tidy(model, conf.int = TRUE) %>%
  mutate(
    term = str_replace_all(term, c(
      "SpeedLimit"                   = "Speed Limit (mph)",
      "log\\(aadt_count \\+ 1\\)"   = "Log(AADT)",
      "urban_ruralUrban"             = "Urban (ref: Rural)",
      "alcoholYes"                   = "Alcohol Involved",

      "lightingDark - Not Lighted"   = "Dark - Not Lighted",
      "lightingDark - Lighted"       = "Dark - Lighted",
      "lightingDawn/Dusk"            = "Lighting: Dawn/Dusk",
      "lightingUnknown"              = "Lighting: Unknown",
      "SpeedLimit:urban_ruralUrban"  = "Speed Limit x Urban"
    )),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      p.value < 0.1   ~ ".",
      TRUE            ~ ""
    )
  )

model_fit <- glance(model)

# Formatted regression results table
model_tidy %>%
  select(Term         = term,
         Estimate     = estimate,
         `Std. Error` = std.error,
         `t-stat`     = statistic,
         `p-value`    = p.value,
         `Sig.`       = sig) %>%
  mutate(across(where(is.numeric), ~round(.x, 4))) %>%
  kable(caption = "Table 3: OLS Regression — Predictors of Fatal Crash Severity",
        align    = c("l", "r", "r", "r", "r", "c"),
        booktabs = TRUE) %>%
  kable_styling(full_width = FALSE,
                latex_options = c("hold_position", "striped")) %>%
  footnote(general = paste0(
    "*** p<0.001  ** p<0.01  * p<0.05  . p<0.1 | ",
    "Ref: Rural, Daylight, No alcohol | ",
    "R2 = ", round(model_fit$r.squared, 3),
    " | n = ", scales::comma(model_fit$nobs)
  ),
  escape = FALSE)
```

```{r plot-coef, fig.cap="Figure 6: Regression coefficient plot with 95% confidence intervals. Red = positive association with fatalities; blue = negative."}
model_tidy %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x     = estimate,
             y     = reorder(term, estimate),
             xmin  = conf.low,
             xmax  = conf.high,
             color = estimate > 0)) +
  geom_pointrange(linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_manual(
    values = c("TRUE" = "#d62728", "FALSE" = "#3182bd"),
    guide  = "none"
  ) +
  labs(
    title    = "Regression Coefficients: Predictors of Crash Fatalities",
    subtitle = paste0("OLS with 95% Confidence Intervals  |  R² = ",
                      round(model_fit$r.squared, 3),
                      "  |  n = ", scales::comma(model_fit$nobs)),
    x        = "Estimated Effect on Fatalities per Crash",
    y        = NULL,
    caption  = "Sources: NHTSA FARS; NCDOT Speed Limits & AADT"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40", size = 10)
  )
```

Table 3 and Figure 6 present the regression results. Interpreting the coefficients relative to the mean of approximately 1.086 fatalities per crash, the key findings are as follows.

**Alcohol involvement** is the strongest and only statistically significant predictor, associated with approximately 2.9% more fatalities per crash compared to non-alcohol crashes. This is consistent with the broader traffic safety literature and underscores impaired driving as the most actionable target for policy intervention.

**Posted speed limit** shows a positive but modest association: a 10 mph increase in posted speed limit is associated with approximately 0.7% more fatalities per crash, after controlling for traffic volume, urban/rural setting, and lighting. While the direction is consistent with the Power Model predictions from the literature, the effect is small in magnitude.

**Traffic volume (AADT)** shows a positive association, with higher-volume roads associated with roughly 1.5% more fatalities per crash moving from low- to high-volume roads. This may reflect multi-vehicle crash dynamics on busier corridors.

**Urban roads** are associated with approximately 3.6% fewer fatalities per crash than rural roads, consistent with faster emergency response times and lower posted speeds in urban areas. The interaction term between speed limit and urban/rural setting is not meaningful, suggesting the speed-fatality relationship does not differ significantly by setting in this dataset.

The model R² of `r round(model_fit$r.squared, 3)` indicates that the included variables explain approximately `r round(model_fit$r.squared*100, 1)`% of the variation in fatalities per crash. This is modest but expected: individual crash outcomes depend heavily on unobserved factors such as seatbelt use, vehicle type, exact point of impact, and road geometry. The low R² does not invalidate the directional findings but does suggest that future research incorporating additional variables would produce more precise estimates.
# Policy Implications

The findings of this analysis have direct relevance for NCDOT and the NC Governor's Highway Safety Program under the Vision Zero initiative.

**If speed limits are a significant positive predictor of fatality risk**, this provides NC-specific quantitative evidence supporting review and targeted reduction of speed limits on high-risk corridors. Priority should be given to rural two-lane highways with posted limits of 55 mph or above, where fatal crashes concentrate and emergency response times are longest. Even modest reductions — 55 to 50 mph — produce meaningful fatality reductions under the Power Model framework.

**If traffic volume (AADT) is the stronger predictor**, the policy focus should shift toward infrastructure improvements on high-volume roads: median barriers, rumble strips, improved lighting, and dedicated turning lanes to reduce crash severity where exposure is highest.

**If the rural interaction term is significant**, NCDOT should explicitly incorporate rural context — road geometry, shoulder width, and emergency response time — into speed limit setting procedures, rather than relying solely on traffic volume and design speed.

A key limitation is the use of fatal crashes only, without a denominator of all crashes, making it difficult to compute true fatality rates per vehicle mile traveled. Future research should incorporate NC's full crash database to model the probability of a fatal outcome given any crash. A spatial fixed-effects model accounting for county-level unobserved heterogeneity (enforcement levels, road maintenance quality) would further strengthen causal inference.

# Conclusion

This report examined the association between posted speed limits, traffic volume, and traffic fatality severity across North Carolina using five years of FARS crash data spatially joined to NCDOT speed limit and AADT datasets. The spatial analysis reveals that fatal crashes on high-speed roads concentrate along rural corridors in the eastern coastal plain and Piedmont, while urban crashes cluster near major metro areas at lower speed limits.

The regression results indicate that alcohol involvement is the strongest predictor of crash severity, associated with roughly 2.9% more fatalities per crash. Speed limits and traffic volume show positive but modest associations with fatality severity — a 10 mph speed limit increase corresponds to approximately 0.7% more fatalities per crash, while moving from low- to high-volume roads is associated with roughly 1.5% more fatalities per crash. Urban roads average 3.6% fewer fatalities than rural roads.

Taken together, these findings suggest that North Carolina's Vision Zero strategy should prioritize impaired driving enforcement as the most evidence-backed intervention, while also targeting rural high-speed corridors for speed limit review and investing in safety infrastructure on high-volume roads. Future research incorporating the full NC crash database, seatbelt use data, and road geometry characteristics would substantially strengthen these findings.
